A partisan affair: Mapping patronage in municipal bureaucracies of Brazil

### 2021-01-18 paper/draft.tex
finalize characterization of municipalities

Partisanship in Brazil is voluntary and widespread, with over 11 percent of registered voters affiliated to a party \citep{speck2015estudo}\footnote{For context, in most OECD countries party registration does not exceed 5 percent of the electorate. See \citet{biezen2014decline}.}. Registration occurs in the following sequence: a voter reports to a local party office, and the party officials then register the voter officially through the \emph{Tribunal Regional Eleitoral} (Regional Electoral Office). This registration is then centralized by the \emph{Tribunal Superior Eleitoral} (Supreme Electoral Office) and updated accordingly. If there are overlapping registrations, former ones become annulled and are reported as irregular to local party officials. Each voter is therefore only allowed to register for a single party, without any ceilings or floors regarding the duration of this affiliation.

One thing is clear: party ties at the electoral level are durable, with most voters remaining affiliated to a single party for their entire life, as highlighted by figure \emph{FIGURE}. Noting that party affiliations are registered at the municipal level, this empirical fact aligns with qualitative evidence provided by \citet{palmeira1995comicios}, who notes that parties at the local level constitute \emph{grupos pol\'{i}ticos} (political factions) with well-defined boundaries and power disputes. This relative stability of party ties at the electorate level for the minority of voters who are registered with a parties suggests a distinct dynamic tying an elite group of party members to the city hall.

@@

>>>>

\subsection{Brazilian municipal government}

Brazil is a federal republic comprised of 26 states and over 5500 municipalities. Each municipality is headed by a mayor and city councilors (\emph{vereadores}). Local officials are elected for a four-year term, with reelection. Each local election takes place at the same time, in October, with the new administration taking office in January of the following year. With decentralization embedded in the enactment of a new constitution in 1988, local municipalities were given large autonomy with respect to public services such as education and healthcare, as well as building an administrative infrastructure to oversee its daily operations \citep{souza2017modernizaccao}.

Municipal budgets are financed with a mix of federal transfers and local revenues. Smaller municipalities tend to rely on federal transfers, which are subject to oversight by higher levels of government, either federal or state, but in practice are under large discretion by municipal governments. The past two decades have seen a growing share of expenditures social service concentrated by municipalities, which has led to expansion in access to and the quality of local services \citep{arretche2015trajetorias}. Additionally, laws and regulations concerning local economies and society are largely autonomous and instituted by the mayor's office, subject to revision and approval by the local city council \citep{brelaz2013processo}.

Finally, it municipalities are in their vast majority quite small, with approximately 90 percent of municipalities having a population of less than 50 thousand people. In contrast, the 27 state capitals (including Bras\'{i}lia) concentrate over 23 percent of the Brazilian population.\footnote{See coverage \href{https://www.google.com/search?q=tamanho+municipio+brasil+poulacao&source=lmns&hl=en&sa=X&ved=2ahUKEwjyiLjLh6buAhXFn-AKHbzEBxAQ_AUoAHoECAEQAA}{here}.} This largely uneven concentration of the Brazilian population across its municipal governments, as well as the prevalence of small, poorer municipalities, has led observants to conclude that municipal politics is often characterized by clientelism, with local political elites concentrating power through the strategic use of public resources \citep{leal2012coronelismo}.

@@

Partisanship in Brazil is voluntary and widespread, with over 11 percent of registered voters affiliated to a party \citep{speck2015estudo}\footnote{For context, in most OECD countries party registration does not exceed 5 percent of the electorate. See \citet{biezen2014decline}.}. Registration occurs in the following sequence: a voter reports to a municipal party office, and the party officials then register the voter officially through the \emph{Tribunal Regional Eleitoral} (Regional Electoral Office). This registration is then collected and centralized by the \emph{Tribunal Superior Eleitoral} (Supreme Electoral Office) and updated accordingly. If there are overlapping registrations, former ones become annulled and are reported as irregular to local party officials. Each voter is therefore only allowed to register for a single party, without any ceilings or floors regarding the duration of this affiliation.

One thing is clear: party ties at the electoral level are durable, with most voters remaining affiliated to a single party for their entire life, as highlighted by figure \emph{FIGURE}. Noting that party affiliations are registered at the municipal level, this empirical fact aligns with qualitative evidence provided by \citet{palmeira1995comicios}, who notes that parties at the local level constitute political factions (\emph{grupos pol\'{i}ticos} ) with well-defined boundaries and power disputes. This relative stability of party ties at the electorate level for the minority of voters who are registered with a parties suggests a distinct dynamic tying an elite group of party members to the city hall.

\section{Data}
